Schema Change: Challenge Problems

### 2024-05-15 paper.tex
the paragraph (rough)

\usepackage{inputenx}
\usepackage[backend=biber]{biblatex}
%\usepackage{natbib}
\addbibresource{paper.bib}

@@

@@

In the simplest cases, divergence can last only for a brief moments in between zero-downtime deployment steps. But divergence can also occur across longer timescales, especially in situations where the different parts of a system are not all centrally controlled. For example, a public web API may need to support applications using old versions of the API for months or years, since it is not realistic to force all consumers to upgrade immediately. Divergence can even be indefinite—\citet{Cambria} addresses collaborative applications where different users may want to collaborate on shared data through different tools using different schemas, with no intention of ever converging.

@@

The Extract Entity challenge for local-first software is similar to the live programming challenge except that the context is an application that offers similar functionality with peer-to-peer collaboration. In this context the schema change does not need to be performed interactively — a developer can take some time to build, test, and package a solution, perhaps using an API or DSL. The hard part comes when the schema change is to be deployed across all the replicas. This deployment must synchronize code and data upgrades (or decouple them as in Cambria~\cite{Cambria}). The deployment must also deal with migrating ``in flight'' operations so that all replicas converge on the same state without data loss, and ideally without centralized coordination. A radical approach could try to incorporate schema change operations into the underlying datastore/CRDT itself and integrate code as well, but layered solutions are welcome too.

@@

Durable computing: pins runtime data frames to code versions. Doesn’t handle problem of global data schema change.
https://restate.dev/blog/solving-durable-executions-immutability-problem/
https://restate.dev/blog/code-that-sleeps-for-a-month/

@@

state opened
  isClosed := false
  on close => closed
\end{lstlisting}
\end{minipage}
\hspace*{2pt}\vline
\begin{minipage}[t]{0.4\textwidth}
\begin{lstlisting}[language=java,morekeywords={on}]
  class Machine
    name: string
    states: State*
    vars: Decl*

@@

The space of possible (well-formed\footnote{The abstract syntax schema is not expressive enough to define all static invariants of the language, such as: states must by unique by name, references variables in actions must be declared, etc.}) changes to a state machine are summarized in Table~\ref{TBL:changes}.
The first column indicates the change category, the second the affected kind of object, and third a short description of how to deal with the respective change category. Some notes about this table:
\begin{itemize}
\item all rows with ``simple'' in the third column only require a quiescent~\cite{Tranquility} point in the interpreter loop to update the run-time structure (e.g., as shown in Figure~\ref{FIG:doorsRuntime}).
\item removing a state, however, is only \textit{structurally} simple, since removal is easy, but special care needs is required if the subject of removal is the current state. In this case, some heuristic is needed, such as: reject the edit, point the current state to the initial state, or some other strategy (e.g., the nearest state, previous state etc.)
\item everywhere a row mentions ``migrate class'' in the third column, data migration is needed: the run-time objects conforming to the old class must be transformed to instances of the updated class, similar to how Smalltalk migrates objects using \lstinline{become:}.
\item removing a transition or changing the event of a transition potentially has to deal with pending events (e.g., in an event queue) expecting such transitions, since such events are now potentially stale. Strategies to deal with this situation include: simply dropping the events, or requiring that such edits cannot be patched at run time when there are pending events.
\item the type change edit of variable requires a strategy for the current value: either discard and reinitialize, or perform value conversion. For instance, if the type change is from boolean to integer, then true could become 1, and false could become 0.
\item note finally that rename variable is mentioned explicitly as a change: this makes it possible to preserve the run-time value of the variable.
\end{itemize}

@@

\subsection{Goals}
As above, the key requirement from the live programming perspective is to “do minimal harm”. In this case, this results in the following goals:

@@

Our goal is to preserve the ability for actors in the system to read and write to a shared todo list in either the old or new schema—for example, they should be able to write to either the scalar assignee field or to the list of assignees.

>>>>

\usepackage{inputenx}
\usepackage[backend=biber]{biblatex}
\addbibresource{paper.bib}

@@

\textcolor{red}{[TVDS] Super rough attempt at the paragraph:}

Modern day software development is collaborative. Single user programming does not exist anymore.
Just as document editing is becoming more collaborative through online systems like pioneered by Google Docs,
programming would benefit form online collaboration. In a sense this is live programming++: whereas traditional
live programming systems assume a single developer editing code and see their changes immediately reflected
in the running system, online collaborative programming is the same but for multiple programmers at the same time.
The key challenge here, we posit, the *schema change*. We take inspiration from database migration,
model evolution, and version control, and identify benchmark problems. This is already evident in live programming:
editing a program, consisting of type definitions defining the structure run-time state and (transient) procedures, functions
etc. operating on that state, requires run-time state to be migrated, after the programmer edits their program.
Adding the dimension of multiple programs brings us into the domain local-first, and decentralized version control, where 
copies are created and edited, which eventually have to be reconciled again (e.g., through merging). 
Whereas schema evolution and its associated migration problems has seen a lot of attention in the database world,
the model-driven world, and the bidirectional transformation (BX) communities, the techniques are still too low-level,
too manual, and too much focused on a single user perspectives. By identifying the similarities between analogous
problems in live, multi-user programming systems, we hope to stimulate further research on more comprehensive, automatic, 
high-level, and online solutions. So that the benefits of live programming can be enjoyed in a global and distributed setting
involving multiple programmers at the same time.

@@

In the simplest cases, divergence can last only for a brief moments in between zero-downtime deployment steps. But divergence can also occur across longer timescales, especially in situations where the different parts of a system are not all centrally controlled. For example, a public web API may need to support applications using old versions of the API for months or years, since it is not realistic to force all consumers to upgrade immediately. Divergence can even be indefinite -- \citet{Cambria} addresses collaborative applications where different users may want to collaborate on shared data through different tools using different schemas, with no intention of ever converging.

@@

The Extract Entity challenge for local-first software is similar to the live programming challenge except that the context is an application that offers similar functionality with peer-to-peer collaboration. In this context the schema change does not need to be performed interactively -- a developer can take some time to build, test, and package a solution, perhaps using an API or DSL. The hard part comes when the schema change is to be deployed across all the replicas. This deployment must synchronize code and data upgrades (or decouple them as in Cambria~\cite{Cambria}). The deployment must also deal with migrating ``in flight'' operations so that all replicas converge on the same state without data loss, and ideally without centralized coordination. A radical approach could try to incorporate schema change operations into the underlying datastore/CRDT itself and integrate code as well, but layered solutions are welcome too.

@@

Durable computing: pins runtime data frames to code versions. Doesn't handle problem of global data schema change.
https://restate.dev/blog/solving-durable-executions-immutability-problem/
https://restate.dev/blog/code-that-sleeps-for-a-month/

@@

state opened
  isClosed := false
  on close => closed
\end{lstlisting}
\end{minipage}
\hspace*{5pt}\vline
\begin{minipage}[t]{0.4\textwidth}
\begin{lstlisting}[language=java,morekeywords={on}]
  class Machine
    name: string
    states: State*
    vars: Decl*

@@

The space of possible (well-formed\footnote{The abstract syntax schema is not expressive enough to define all static invariants of the language, such as: states must by unique by name, references variables in actions must be declared, etc.}) changes to a state machine are summarized in Table~\ref{TBL:changes}.
The first column indicates the change category, the second the affected kind of object, and third a short description of how to deal with the respective change category. Some notes about this table:
\begin{itemize}
\item all rows with ``simple'' in the third column only require a quiescent~\cite{Tranquility} point in the interpreter loop to update the run-time structure (e.g., as shown in Figure~\ref{FIG:doorsRuntime}).
\item removing a state, however, is only \textit{structurally} simple, since removal is easy, but special care is required if the subject of removal is the current state. In this case, some heuristic is needed, such as: reject the edit, point the current state to the initial state, or some other strategy (e.g., the nearest state, previous state etc.)
\item everywhere a row mentions ``migrate class'' in the third column, data migration is needed: the run-time objects conforming to the old class must be transformed to instances of the updated class, similar to how Smalltalk migrates objects using \lstinline{become:}.
\item removing a transition or changing the event of a transition potentially has to deal with pending events (e.g., in an event queue) expecting such transitions, since such events are now potentially stale. Strategies to deal with this situation include: simply dropping the events, or requiring that such edits cannot be patched at run time when there are pending events.
\item the type change edit of variable requires a strategy for the current value: either discard and reinitialize, or perform value conversion. For instance, if the type change is from boolean to integer, then true could become 1, and false could become 0.
\item note finally that rename variable is mentioned explicitly as a change: this makes it possible to preserve the run-time value of the variable.
\end{itemize}

@@

\subsection{Goals}
As above, the key requirement from the live programming perspective is to ``do minimal harm''. In this case, this results in the following goals:

@@

Our goal is to preserve the ability for actors in the system to read and write to a shared todo list in either the old or new schema -- for example, they should be able to write to either the scalar assignee field or to the list of assignees.
